---
title: "NOAA Events damage exploration"
author: "Paulo Medina"
date: "2025-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This report is built based on the **Storm Data** collected and disseminated by the [National Oceanographic and Atmospheric Association](https://www.noaa.gov/). The aim of the report is explore this data set to know which are the most damaging atmospheric events.

The report presents the data processing, as well as the results.

## Data Processing

Below follows a thorough description of the data preparation.

The libraries used for this work are:

* tidyverse: for data prep.
* here: for dir management.
* egg: for plot layouts.

### Data Load

The data is provided as a `csv.bzip` file. To access it, I created a temporary file to store the raw data, then proceeded to read into the temp file with `data.table`.

All the temporary files are then dropped.

```{r libraries, message=FALSE,warning=FALSE, cache=TRUE}
library(tidyverse)
library(here)
library(egg)
```

```{r data load, cache = TRUE}
#Load data

if(!dir.exists("./data")){dir.create(here("./data"))}#Set-up directory for work and data

#Download data

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
temp <- tempfile() #Temporary variable to save zip file
download.file(url,temp)

```

The data is then read.

```{r data read, cache = TRUE}
repdata <- read.table(temp, sep = ",", header = TRUE)

unlink(temp); rm(temp, url) #Delete temporary files

str(repdata)

```

### Subsetting data

Given the size of the data, I will subset the dataset to hold the important variables only, and then group these by event type.

```{r subsetting, cache=TRUE}
eventype <- repdata %>% 
        group_by(EVTYPE) %>%
        summarise(fatalities = sum(FATALITIES), 
                  mag = sum(MAG),
                  injuries = sum(INJURIES),
                  propmg = sum(PROPDMG))

head(eventype,5)
str(eventype)
```

With the relevant data, I will plot the top 20 events by fatalitites and injuries to see which are the most harmful.

### Data exploration

Let's look at the top events in general. The `top_n` function will retrieve the top events.

```{r data exploration, cache=TRUE}
top_n(eventype, 20)
```

Seems that **Flash Flooding** is the top event, with the highest aggregated values for all 4 categories.

With data ready, we will be able to present the results.

## Results

The data indicates that the most harmful events for human life are the **tornadoes**. As seen below, the tornadoes are at the top for most fatalities and injuries

```{r fatalities plot, fig.height=6,fig.width= 10, cache=TRUE}
fat <- eventype %>% 
        top_n(20, fatalities) %>%
        ggplot() + 
        geom_bar(aes(x=fatalities, 
                     y=reorder(EVTYPE,+fatalities)),
                 stat="identity") +
        labs(title="Fatalities",x="",y="")

inj <- eventype %>% 
        top_n(20, injuries) %>%
        ggplot() + 
        geom_bar(aes(x=injuries, 
                     y=reorder(EVTYPE,+injuries)),
                 stat="identity") +
        labs(title="Injuries",x="",y="")

# Plot code:

ggarrange(fat,inj, ncol = 2)

```

In terms of material damage

```{r material damage plot, fig.height=6, fig.width=10, cache=TRUE}
eventype %>% 
        top_n(20, propmg) %>%
        ggplot() + 
        geom_bar(aes(x=propmg, 
                     y=reorder(EVTYPE,+propmg)),
                 stat="identity") +
        labs(title="Property Damage",x="",y="")

```

With this, we can conclude that *Tornadoes* are the most harmful events accross the United States.

